\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}

\title{Core ML: Modular Long-Context Sequence Modeling}
\author{Ansh}
\date{SAiDL Summer Assignment 2026}

\begin{document}
\maketitle

\section{Overview}

This report studies long-context sequence modeling on WikiText-2 using a modular
Transformer language model. The codebase allows independent swaps of attention
mechanisms, positional encodings, and convolution-attention hybrid blocks.

\section{Baseline}

\subsection{Architecture}

The baseline is a pre-normalized causal Transformer language model with tied
token embedding and language-model head weights. The default configuration uses
a context length of 1024 tokens, embedding dimension 512, 8 heads, 6 layers, and
a feed-forward dimension of 2048.

\subsection{Training Setup}

\begin{itemize}
    \item Dataset: WikiText-2 raw.
    \item Tokenizer: GPT-2 tokenizer.
    \item Optimizer: AdamW.
    \item Model width/depth: 512 hidden dimension, 8 attention heads, 6 Transformer layers, and feed-forward dimension 2048.
    \item Sequence length: 1024 tokens for the initial baseline and attention-variant experiments.
    \item Batch size: 4.
    \item Logged metrics: training loss, validation loss, validation perplexity,
    peak GPU memory, throughput, and elapsed time.
\end{itemize}

\begin{table}[h]
\centering
\begin{tabular}{lrrrrr}
\toprule
Model & Context & Val Loss & Val PPL & Val Tokens/sec & Train Peak Mem (MB) \\
\midrule
Standard + Sinusoidal & 1024 & 5.862 & 351.41 & 29636.66 & 4182.75 \\
\bottomrule
\end{tabular}
\caption{Baseline results.}
\end{table}

The standard Transformer with sinusoidal positional encoding serves as the
reference point for all later comparisons. At step 8500 it reaches a validation
loss of 5.862 and validation perplexity of 351.41 while maintaining high
throughput and moderate memory use. This baseline is strong enough to act as a
meaningful control but still leaves ample room for architectural improvement.

\section{Attention Variants}

The implemented attention variants are standard causal self-attention, sliding
window attention, linear kernelized attention, multi-query attention, and
grouped-query attention.

\begin{table}[h]
\centering
\begin{tabular}{lrrrrrr}
\toprule
Attention & Context & Step & Val Loss & Val PPL & Val Tokens/sec & Train Peak Mem (MB) \\
\midrule
Standard & 1024 & 8500 & 5.862 & 351.41 & 29636.66 & 4182.75 \\
Sliding Window & 1024 & 5500 & 6.052 & 424.96 & 28301.10 & 4188.65 \\
Linear & 1024 & 8500 & 5.856 & 349.49 & 23573.86 & 4981.47 \\
GQA & 1024 & 8500 & 5.810 & 333.78 & 29830.34 & 4155.67 \\
\bottomrule
\end{tabular}
\caption{Attention ablation results.}
\end{table}

Among the attention variants evaluated so far, grouped-query attention (GQA)
performs best. It achieves the lowest validation loss and validation perplexity
while also remaining efficient in both throughput and memory. Linear attention
slightly improves over the standard baseline in validation perplexity, but this
comes with noticeably higher memory consumption and lower throughput. The
sliding-window model performs worst in the current results, although that run
was logged at a smaller training budget than the others and should therefore be
treated as a preliminary comparison rather than a final conclusion.

\section{Positional Encodings}

The implemented positional choices are sinusoidal positional encodings, rotary
positional embeddings, ALiBi, and learned relative position bias. The planned
extrapolation test trains with $L_{\mathrm{train}}=512$ and evaluates at
$L_{\mathrm{test}}\in\{512,1024,2048\}$.

\begin{table}[h]
\centering
\begin{tabular}{lrrrrr}
\toprule
Position Method & Attention & Context & Step & Val PPL & Notes \\
\midrule
Sinusoidal & GQA & 1024 & 8500 & 333.78 & Fully completed run \\
RoPE & GQA & 1024 & 6500 & 150.28 & Best logged validation checkpoint \\
RoPE (final) & GQA & 1024 & 8500 & 152.73 & Completed rerun, slight late-stage degradation \\
ALiBi & -- & -- & -- & -- & Pending \\
Relative Bias & -- & -- & -- & -- & Pending \\
\bottomrule
\end{tabular}
\caption{Current positional encoding results.}
\end{table}

RoPE is the strongest positional encoding tested so far in the current setup.
The completed GQA + RoPE rerun achieved its best logged validation performance
at step 6500, reaching validation loss 5.012 and validation perplexity 150.28.
After this point the validation curve drifted upward slightly, and the final
checkpoint at step 8500 ended at validation loss 5.029 and validation
perplexity 152.73. Even this final checkpoint remains dramatically better than
the GQA + sinusoidal baseline, confirming that RoPE provides a substantial
improvement for long-context language modeling in this model family. ALiBi and
relative positional bias are implemented but had not yet been fully evaluated at
the time of writing.

\section{Convolution-Attention Hybrids}

Two hybrid designs are implemented: a depthwise convolution branch before each
attention layer, and a gated convolutional feed-forward block. These are tested
with the best attention and positional encoding from the earlier ablations.

\begin{table}[h]
\centering
\begin{tabular}{lrrrr}
\toprule
Block Type & Val PPL & Tokens/sec & Peak Memory MB & Notes \\
\midrule
Standard & 152.73 (RoPE run) & 31379.15 & 4155.07 & Best current non-hybrid model \\
Conv Before Attention & -- & -- & -- & Pending \\
Gated Conv FFN & -- & -- & -- & Pending \\
\bottomrule
\end{tabular}
\caption{Hybrid architecture results.}
\end{table}

At the current stage of the project, the convolution-augmented variants are
implemented in code but have not yet been benchmarked thoroughly. The strongest
non-hybrid configuration so far is GQA with RoPE and a standard Transformer
block, which provides the natural starting point for the hybrid experiments.

\section{Analysis}

The first major conclusion from the experiments is that not all efficient
attention variants are equally useful in this setting. Sliding-window attention
did not improve either quality or efficiency enough to justify its use at the
tested budget. Linear attention improved validation perplexity slightly over the
standard baseline but required substantially more training memory and reduced
throughput, which weakens its appeal as a practical replacement.

GQA, in contrast, improved both language modeling quality and efficiency. It
reduced validation perplexity from 351.41 to 333.78 while slightly lowering peak
training memory and maintaining very similar throughput. This makes GQA the most
attractive attention choice among the tested variants.

The positional encoding experiments point even more strongly toward the
importance of architecture choices for long-context behavior. RoPE produced a
very large improvement when paired with GQA, reducing validation perplexity from
333.78 for GQA + sinusoidal to a best logged value of 150.28. The validation
curve indicates that the model performs best around step 6500 and then degrades
slightly afterward, suggesting mild late-stage overfitting rather than
instability. This is still a clear win for RoPE and makes it the strongest
positional encoding result in the project so far.

Overall, the experiments suggest a useful hierarchy: GQA is the best attention
mechanism tested so far, and RoPE is the most promising positional encoding.
This already gives a strong candidate model for the next stages of the project
and for later hybrid convolution experiments.

\section{Conclusion}

The baseline standard Transformer with sinusoidal positional encoding establishes
a reasonable starting point, but it is clearly not the strongest model in the
current study. GQA emerges as the best attention variant, outperforming
standard, sliding-window, and linear attention on validation perplexity while
remaining efficient in both memory and throughput. RoPE further improves results
substantially when combined with GQA, making GQA + RoPE the strongest model
observed so far. The best logged RoPE checkpoint occurs at step 6500, where the
model reaches validation perplexity 150.28, while the final completed run at
step 8500 remains close at 152.73.

The main lesson from the ablation study is that long-context modeling quality in
this setup depends strongly on choosing the right attention and positional
components rather than only scaling the baseline Transformer. Future work in the
same codebase should complete the remaining positional and convolution-hybrid
ablations, then carry the best-performing configuration forward as the final Core
ML submission model.

\end{document}
