\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}

\title{Diffusion: Regionally-Adaptive Cyclic Diffusion}
\author{Ansh}
\date{SAiDL Summer Assignment 2026}

\begin{document}
\maketitle

\section{Overview}

This report studies Regionally-Adaptive Cyclic Diffusion (RACD) for landscape
image generation. The assignment specifies a landscape dataset but does not
provide one directly, so the experiments use a public Kaggle landscape image
dataset resized to $256 \times 256$. The implementation uses latent diffusion:
images are encoded with a frozen pretrained VAE, and a compact DiT-style
Transformer is trained to predict diffusion noise in latent space.

\section{Method}

\subsection{Baseline Latent DiT}

The baseline model is a compute-aware latent DiT trained on frozen VAE latents.
To fit Kaggle free-tier constraints, the initial model uses a smaller
configuration than DiT-B/8: hidden size 384, 6 layers, 6 attention heads, and
patch size 4 in the $32 \times 32$ latent grid. This gives 64 latent tokens per
image, which preserves more spatial structure for later region-adaptive
refinement.

\subsection{Global Cyclic Refinement}

After baseline sampling reaches $z_0$, global cyclic refinement re-injects noise
up to timestep $t=200$ and denoises back to $t=0$. This applies extra compute to
the entire latent and serves as the main non-adaptive refinement baseline.

\subsection{Difficulty Predictor}

The difficulty predictor is trained after the DiT is frozen. For each training
image, a noisy latent $z_t$ is generated at an intermediate timestep. The frozen
DiT predicts the noise, which is converted into an intermediate $\hat{z}_0$
estimate. The per-location squared error between $\hat{z}_0$ and the clean VAE
latent becomes the difficulty supervision target. A small CNN maps intermediate
DiT token features to a $32 \times 32$ difficulty map.

\subsection{RACD}

At sampling time, the difficulty predictor estimates which latent regions should
receive refinement. A threshold $\tau$ converts the difficulty map into a binary
mask. Noise is re-injected only in masked regions, then the model denoises from
$t=200$ back to $t=0$. The final latent combines refined masked regions with
unmasked baseline regions.

\section{Experiments}

\begin{table}[h]
\centering
\begin{tabular}{lrrrr}
\toprule
Method & FID $\downarrow$ & CMMD $\downarrow$ & Time (s/img) & Refinement Fraction \\
\midrule
Baseline DiT & -- & -- & -- & 0.00 \\
Global Cyclic & -- & -- & -- & 1.00 \\
RACD $\tau=0.3$ & -- & -- & -- & -- \\
RACD $\tau=0.5$ & -- & -- & -- & -- \\
RACD $\tau=0.7$ & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\caption{Diffusion evaluation table to be filled after Kaggle sampling runs.}
\end{table}

\section{Analysis}

The primary quantities of interest are whether global cyclic refinement improves
sample quality at a clear compute cost, and whether RACD can recover most of the
quality gain while refining only a fraction of latent regions. The final report
will compare FID, CMMD, generation time, and visual grids for baseline, global
cyclic, and RACD threshold sweeps.

\section{Conclusion}

This section will be completed after the baseline DiT, global cyclic refinement,
and RACD experiments finish.

\end{document}
